---
title: "Assignment 5B – ELO Calculations Codebase"
author: "Muhammad Suffyan Khan"
format: html
editor: visual
date: "2026-03-01"
---

### Objective

The objective of this assignment is to compute each player’s **expected tournament score** using the Elo rating model and compare it with their actual score from the Project 1 tournament.

We will then identify:

-   The five players who most **overperformed** relative to expectation\
-   The five players who most **underperformed** relative to expectation

------------------------------------------------------------------------

### Data Source

This analysis builds directly on the structured dataset created in Project 1.

The dataset contains exactly one row per player with the following variables:

-   Player Name\
-   Player State\
-   Total Points (Actual Score)\
-   Pre-Tournament Rating\
-   Average Pre-Tournament Rating of Opponents

This dataset was exported as a CSV file and uploaded to GitHub to ensure reproducibility. The file will be loaded into R using the GitHub raw link.

------------------------------------------------------------------------

### Elo Expected Score Formula

The Elo rating system models the probability that Player A defeats Player B using a logistic function based on rating differences.

The expected score formula used in this assignment is:

$$
E_A = \frac{1}{1 + 10^{(R_B - R_A)/400}}
$$

Where:

-   (R_A) = Player A’s pre-tournament rating\
-   (R_B) = Opponent’s pre-tournament rating\
-   400 = scaling constant in the Elo system

This formulation implies that a 400-point rating difference corresponds to approximately a 10-to-1 expected win ratio.

**Sources:**\
- Glickman, M. E. *A Comprehensive Guide to Chess Ratings*: <https://www.glicko.net/research/acjpaper.pdf>\
- *Elo rating system* (Wikipedia): <https://en.wikipedia.org/wiki/Elo_rating_system>\
- *The Elo Rating System for Chess and Beyond* (YouTube, 2019): <https://www.youtube.com/watch?v=AsYfbmp0To0>

------------------------------------------------------------------------

### Approximation Strategy

In the standard Elo system, expected score is computed separately for each game using each opponent’s rating, and then summed across games.

However, the Project 1 dataset retains only the **average pre-tournament rating of opponents**, rather than the full match level opponent list. Therefore, this analysis approximates expected tournament score by applying the Elo expected score formula using the player’s rating versus their **average opponent rating**, and then scaling by the number of rounds.

The expected per-game score is computed as:

$$
E = \frac{1}{1 + 10^{(\overline{R_{opp}} - R_{player})/400}}
$$

The expected tournament score is then:

$$
\text{Expected Tournament Score} = n \times E
$$

Where:

-   \(n\) = number of rounds (games) in the tournament\
-   ({R\_{opp}}) = average opponent pre-rating

This provides a transparent approximation consistent with the Elo logistic framework given the available Project 1 output.

------------------------------------------------------------------------

### Implementation Steps

The analysis will proceed as follows:

1.  **Load Project 1 dataset**
    -   Load `Project1_Chess_Summary.csv` directly from the GitHub raw link using `read.csv()` to ensure reproducibility.
2.  **Compute expected per-game probability**
    -   For each player, compute expected per-game score using the Elo formula and the player’s pre-rating versus the average opponent pre-rating.
3.  **Multiply by number of rounds**
    -   Since the tournament consists of seven rounds, the expected per-game probability will be multiplied by the number of rounds played to estimate the player’s total expected tournament score.
4.  **Compute performance difference**
    -   Compute:

$$
\text{Performance Difference} = \text{Actual Score} - \text{Expected Score}
$$

-   Positive values indicate overperformance relative to rating.
-   Negative values indicate underperformance.

5.  **Rank players**
    -   Sort players by performance difference.
    -   Report the top five overperformers and bottom five underperformers.

------------------------------------------------------------------------

### Final Output

The final output will include:

-   Player Name\
-   Pre-Tournament Rating\
-   Actual Score\
-   Expected Score (approximate)\
-   Performance Difference

Players will be ranked to identify:

-   Top 5 Overperformers\
-   Top 5 Underperformers

This approach builds directly upon the structured output from Project 1 and applies the Elo rating model to evaluate tournament performance relative to expectation.

------------------------------------------------------------------------

## Codebase

```{r,message: false ,warning: false}

library(tidyverse)
library(knitr)
```

## 1. Load Project 1 dataset

This analysis uses the cleaned Project 1 summary dataset (one row per player) containing:

-   Player Name

-   Player State

-   Total Points (Actual Score)

-   Pre-Tournament Rating

-   Average Pre-Tournament Rating of Opponents

```{r}
chess <- read.csv(
  "https://raw.githubusercontent.com/suffyankhan77/Assignment5B-DATA-607/main/Project1_Chess_Summary.csv"
)

chess <- as_tibble(chess)

glimpse(chess)
colnames(chess)
```

## 2. Elo expected score formula (per-game)

We use the standard Elo expected score formula:

$$
E_A = \frac{1}{1 + 10^{(R_B - R_A)/400}}
$$

**This follows the standard Elo logistic model (see Approach section for detailed explanation and sources).**

## 3. Compute expected tournament score (approximation)

Because the Project 1 output retains only the average opponent pre-rating (not the full opponent list), I approximated the expected tournament score by computing expected per-game score versus the average opponent rating, then multiplying by the number of rounds played.

In this tournament there are 7 rounds, so we use `n_rounds = 7`.

```{r}
n_rounds <- 7

elo_expected <- function(r_player, r_opp_avg) {
  1 / (1 + 10^((r_opp_avg - r_player) / 400))
}

results <- chess %>%
  mutate(
    expected_per_game = elo_expected(Pre_Rating, Average_Opponent_Pre_Rating),
    expected_score = n_rounds * expected_per_game,
    actual_score = Total_Points,
    performance_diff = actual_score - expected_score
  )

results %>%
  select(Player_Name, State, Pre_Rating, Average_Opponent_Pre_Rating, actual_score, expected_score, performance_diff) %>%
  arrange(desc(performance_diff)) %>%
  head(10) %>%
  kable(digits = 3, caption = "Top 10 players by performance difference (Actual - Expected)")
```

## 4. Identify overperformers and underperformers

We rank players by: $$
Performance Difference=Actual Score−Expected Score
$$\

Positive values indicate overperformance\
Negative values indicate underperformance

## Top 5 Overperformers

```{r}
top_5_over <- results %>%
  arrange(desc(performance_diff)) %>%
  select(Player_Name, State, Pre_Rating, actual_score, expected_score, performance_diff) %>%
  slice(1:5)

kable(top_5_over, digits = 3, caption = "Top 5 Overperformers (Actual - Expected)")
```

## Top 5 Underperformers

```{r}
top_5_under <- results %>%
  arrange(performance_diff) %>%
  select(Player_Name, State, Pre_Rating, actual_score, expected_score, performance_diff) %>%
  slice(1:5)

kable(top_5_under, digits = 3, caption = "Top 5 Underperformers (Actual - Expected)")
```

## 5. Visualization

```{r}
ggplot(results, aes(x = performance_diff)) +
  geom_histogram(bins = 15) +
  labs(
    title = "Distribution of Performance Difference (Actual - Expected)",
    x = "Performance Difference",
    y = "Number of Players"
  )
```

## 6. Brief Interpretation

Players with the largest positive performance differences significantly exceeded their expected scores based on pre-tournament ratings and average opponent strength. In several cases, lower-rated players substantially outperformed rating-based expectations.

Conversely, the largest negative differences indicate higher-rated players who scored fewer points than predicted by the Elo model.

Because expected scores were calculated using average opponent ratings (rather than game-by-game calculations), these results represent an approximation of true Elo expectations. However, the ranking clearly highlights players whose tournament performance diverged most from rating-based predictions.
